---
title: "RNA-seq prep"
author: "Daniel Hammarström"
date: "4 desember 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
source("./R/lib.R")
```

# Background and brief method
Thirty-four participants completed a 12-wk training-intervention with each leg allocated to either low- or moderate-volume training (one or three sets of 6-12 repetition maximum). The study was primarily designed to assess the effect of training volume on muscular-hypertrophy and strength. Muscle biopsies were collected prior to and after the intervention as well as prior to and 60-min after the fifth training session. Total RNA was extracted by a standard Trizol protocol and concentrations were determined on a spectrophotometer. RNA-integrity numbers were determined using Bio-Rad Experion Total RNA kit.

# Results

## Training outcomes

Training with multiple-sets per excercise was more efficient than single-set training (fig. 1).
```{r training outcomes, include=TRUE, warning=FALSE, echo=FALSE, message=FALSE}
### Fig 1 ----- Muscle CSA pre-post ###########

source("./R/MR_extractRaw.R");source("./R/functions.R")
include<-read.csv("./data/oneThreeSetLeg.csv", sep=";")

include<-include%>%
  gather(sets,leg, multiple:single)%>%
  mutate(condition=leg)%>%
  dplyr::select(subject, sex, include, condition, sets)


mr<-mr.results%>%
  mutate(condition=leg)%>%
  inner_join(include)%>%
  filter(include=="incl")%>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg)%>%
  spread(timepoint, CSA.avg)%>%
  mutate(csa.change=((post/pre)-1)*100)


summary.mr<-mr%>%
  group_by(sets)%>%
  summarise(n=n(),
            s=sd(csa.change),
            error=qt(0.975, df=n-1)*s/sqrt(n),
            csa.change=mean(csa.change))%>%
  ungroup()%>%
  mutate(sets=factor(sets, levels=c("single", "multiple"), labels=c("Single-set", "Multiple-sets")),
         subject="group")

### Statistics 

# paired t-test percentage change in 

temp<-mr.results%>%
  mutate(condition=leg)%>%
  inner_join(include)%>%
  filter(include=="incl")%>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg)%>%
  spread(timepoint, CSA.avg)%>%
  mutate(csa.change=(post-pre))%>%
  dplyr::select(subject, sex, sets, csa.change)%>%
  spread(sets, csa.change)


mr.ttest<-with(temp, t.test(multiple, single, paired=T))


### plot
fig1a<-mr%>%
  mutate(sets=factor(sets, levels=c("single", "multiple"), labels=c("Single-set", "Multiple-sets")))%>%
  ggplot(aes(sets, csa.change, group=subject))+geom_line()+
  geom_errorbar(data=summary.mr[summary.mr$sets=="Single-set",], aes(ymin=csa.change-error, ymax=csa.change+error),
                position = position_nudge(x=-0.1), width=0)+
  geom_errorbar(data=summary.mr[summary.mr$sets=="Multiple-sets",], aes(ymin=csa.change-error, ymax=csa.change+error),
                position = position_nudge(x=0.1), width=0)+

  geom_point(data=summary.mr[summary.mr$sets=="Single-set",], aes(sets, csa.change), pch=21,fill="white",size=3,
             position = position_nudge(x=-0.1))+
 
  geom_point(data=summary.mr[summary.mr$sets=="Multiple-sets",], aes(sets, csa.change), pch=21,fill="black",size=3,
             position = position_nudge(x=0.1))+
  theme(legend.position=c(.2, .95),
        axis.line.x = element_line(color="black", size = 0.5, lineend="square"),
        axis.line.y = element_line(color="black", size = 0.5, lineend="square"),
        axis.ticks.x=element_line(color="black", size = 0.5),
        axis.ticks.y=element_line(color="black", size = 0.5),
        axis.text=element_text(colour="black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        legend.background=element_blank(),
        legend.key=element_blank(),
        legend.title=element_blank(),
        strip.background=element_blank())+
 xlab("")+ylab("CSA relative change (%)")+
  annotate("text", x=1.5, y=12, label=pval(mr.ttest$p.value), parse=FALSE)


temp.stats<-mr.results%>%
  mutate(condition=leg)%>%
  inner_join(include)%>%
  filter(include=="incl")%>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg)%>%
  spread(timepoint, CSA.avg)%>%
  mutate(csa.change=(post-pre))%>%
  dplyr::select(subject, csa.change, sets)%>%
  spread(sets, csa.change)

r2<-round(cor.test(temp.stats$single, temp.stats$multiple)$estimate^2,3)
p<-pval(cor.test(temp.stats$single, temp.stats$multiple)$p.value)
  
fig1b<-mr.results%>%
  mutate(condition=leg)%>%
  inner_join(include)%>%
  filter(include=="incl")%>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg)%>%
  spread(timepoint, CSA.avg)%>%
  mutate(csa.change=(post-pre))%>%
  dplyr::select(subject, sex, sets, csa.change)%>%
  spread(sets, csa.change)%>%
  mutate(sex=factor(sex, levels=c("female", "male"), labels=c("Female", "Male")))%>%
  ggplot(aes(single, multiple, shape=sex, fill=sex))+
  geom_point(size=3)+
  geom_abline(intercept=0, slope=1)+
  scale_x_continuous(limits=c(-1.2, 10))+scale_y_continuous(limits=c(-1.2, 10))+
  theme(legend.position=c(.2, .95),
        axis.line.x = element_line(color="black", size = 0.5, lineend="square"),
        axis.line.y = element_line(color="black", size = 0.5, lineend="square"),
        axis.ticks.x=element_line(color="black", size = 0.5),
        axis.ticks.y=element_line(color="black", size = 0.5),
        axis.text=element_text(colour="black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        legend.background=element_blank(),
        legend.key=element_blank(),
        legend.title=element_blank(),
        strip.background=element_blank())+
  scale_shape_manual(values=c(24,25))+
  scale_fill_manual(values=c("white", "black"))+
xlab(expression(paste("Single-set CSA absolute change (",cm^2,")")))  +
       ylab(expression(paste("Multiple-sets CSA absolute change (",cm^2,")")))+
  annotate("text", x=6, y=2.5, label=paste("R^2 == ", r2), parse=TRUE)+
  annotate("text", x=6, y=1.8, label=p, parse=TRUE)


g1 <- arrangeGrob(fig1a, top = textGrob("A", x = 0, hjust = -0.8, vjust=0.8, gp = gpar(fontface = "bold", fontsize = 16)))
g2 <- arrangeGrob(fig1b, top = textGrob("B", x = 0, hjust = -0.8, vjust=0.8, gp = gpar(fontface = "bold", fontsize = 16)))

grid.arrange(g1,g2, ncol=2, top="Figure 1. Training outcomes.")
```

## RNA integrity
RNA integrity was compromised in some samples. RNA integrity vs. extraction parameters are shown in figure 2.

```{r RNA integrity, include=TRUE, warning=FALSE, echo=FALSE, message=FALSE, fig.height=4, fig.width=6}

rna_integrity <- read_excel("./data/RNAintegrity.xlsx", sheet = 2)

temp <- data.frame(str_split_fixed(rna_integrity$`Sample Name`, " ", 3))

temp$subject <- paste("FP", temp$X1, sep="")
colnames(temp) <- c("subjnr", "leg", "timepoint", "subject")


RNA_integrity <- cbind(temp[, c(2:4)], rna_integrity[,c(3, 4, 5, 6, 7)])

colnames(RNA_integrity) <- c("leg", "timepoint", "subject","RNA_area",  "RNA_conc", "RNA_ratio", "RQI", "Classification")

rna_subjinfo <- read_excel("./data/RNAintegrity.xlsx", sheet = 1)

RNA<-RNA_integrity%>%
  inner_join(rna_subjinfo[ , c(1,3,4,5,6,7,8, 9)])%>%
  mutate(RQI = as.numeric(RQI))

fig2<-RNA%>%
  ggplot(aes(prot.mrna1, RQI, size=elution.volume*conc, fill=Classification))+geom_point(shape=21)+
  scale_fill_manual(name="RQI classification", values=c("green", "red", "yellow"))+
  scale_size(name="Total RNA\namount")+xlab("Tissue weight")


fig2+ggtitle("Figure 2. RNA integrity and extraction parameters")
```

 

```{r RNA integrity and hypertrophy outcomes, include=TRUE, warning=FALSE, echo=FALSE, message=FALSE, fig.height=4, fig.width=6}

rna_min<-RNA%>%
  group_by(subject)%>%
  summarise(RQI.low = min(RQI))


subj.with.lowrqi<-rna_min%>%
  filter(RQI.low < 7.4)%>%
  nrow()


fig3<-mr%>%
  inner_join(rna_min)%>%
  mutate(RQI.class = ifelse(RQI.low > 7, 
                            yes="Green (Minimum RQI > 7)", 
                            no = ifelse(RQI.low < 7 & RQI.low > 4, yes = "yellow", no = "Red (Minimum RQI < 4)")))%>%
  mutate(sets = factor(sets, levels = c("single", "multiple")))%>%
  ggplot(aes(sets, csa.change, fill=RQI.class, group=subject, shape=sex))+geom_line()+geom_point(size=3)+
  scale_fill_manual(name="RQI classification", values=c("green", "red"))+scale_shape_manual(values = c(21, 22))+
  facet_grid(.~RQI.class)

incl<-mr%>%
  inner_join(rna_min)%>%
  mutate(RQI.class = ifelse(RQI.low > 7, 
                            yes="Green (Minimum RQI > 7)", 
                            no = ifelse(RQI.low < 7 & RQI.low > 4, yes = "yellow", no = "Red (Minimum RQI < 4)")))%>%
  filter(RQI.low > 7)%>%
  dplyr::select(subject, sex, sets, csa.change)%>%
  spread(sets, csa.change)


mr%>%
  inner_join(rna_min)%>%
  mutate(RQI.class = ifelse(RQI.low > 7, 
                            yes="Green (Minimum RQI > 7)", 
                            no = ifelse(RQI.low < 7 & RQI.low > 4, yes = "yellow", no = "Red (Minimum RQI < 4)")))%>%
  mutate(sets = factor(sets, levels = c("single", "multiple")))%>%
  dplyr::select(subject, sex, sets, csa.change, RQI.class) %>%
  spread(sets, csa.change) %>%
  ggplot(aes(single, multiple, fill = RQI.class)) + geom_point(size = 2, shape = 21) + 
  geom_abline(slope = 1, intercept = 0) + 
  scale_fill_manual(values = c("green", "red"))
  
### Calculate the mean error at baseline as an estimate of measurement error
error <- mr.results %>%
  mutate(condition=leg) %>%
  inner_join(read.csv("./data/oneThreeSetLeg.csv", sep=";") %>%
               gather(sets,leg, multiple:single) %>%
               mutate(condition = leg) %>%
               dplyr::select(subject, sex, include, condition, sets)) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  filter(timepoint == "pre") %>%
  group_by(subject) %>%
  summarise(m = mean(CSA.avg), 
            s = sd(CSA.avg)) %>%
  ungroup() %>%
  mutate(error = (s/m) * 100) %>%
  summarise(error = mean(error)) 


benefit <- mr.results %>%
  mutate(condition=leg) %>%
  inner_join(read.csv("./data/oneThreeSetLeg.csv", sep=";") %>%
               gather(sets,leg, multiple:single) %>%
               mutate(condition = leg) %>%
               dplyr::select(subject, sex, include, condition, sets)) %>%
  filter(include=="incl") %>%
  dplyr::select(subject, timepoint, sex, sets, CSA.avg) %>%
  spread(timepoint, CSA.avg) %>%
  mutate(error = as.numeric(error), 
         difference = post - pre) %>%
  inner_join(read_excel("./data/bodycomp_DXA.xlsx") %>%
               inner_join(read.csv("./data/oneThreeSetLeg.csv", sep=";")) %>%
               filter(include =="incl") %>% 
               group_by(subject) %>%
               summarise(weight = mean(weight))) %>% 
  group_by(subject) %>%
  mutate(absolute.error = mean(pre) * (error/100)) %>%
  ungroup() %>%
  dplyr::select(subject, sex, sets, difference, absolute.error) %>%
  spread(sets, difference) %>%
  mutate(benefit = if_else(multiple - single > absolute.error, 1, 0)) %>%
  dplyr::select(subject, sex, benefit) 




group1 <- mr%>%
  inner_join(rna_min) %>%
  inner_join(benefit) %>%
  mutate(RQI.class = ifelse(RQI.low > 7, 
                            yes="Green (Minimum RQI > 7)", 
                            no = ifelse(RQI.low < 7 & RQI.low > 4, yes = "yellow", no = "Red (Minimum RQI < 4)")))%>% 
  mutate(sets = factor(sets, levels = c("single", "multiple")))%>%
  dplyr::select(subject, sex, sets,benefit, csa.change, RQI.class) %>%
  spread(sets, csa.change) %>%
  filter(RQI.class != "Red (Minimum RQI < 4)") %>%
  ggplot(aes(single, multiple, fill = as.factor(benefit))) + geom_point(shape = 21, size = 2) +
   geom_abline(intercept = 0, slope = 1, color = "gray60") + 
  geom_abline(intercept = 2.4, slope = 1, color = "gray20")  +
   scale_fill_manual(values = c("blue", "red"), name = "Benefit of\nmultiple-sets")
  
temp <- mr%>%
  inner_join(rna_min) %>%
  inner_join(benefit) %>%
  mutate(RQI.class = ifelse(RQI.low > 7, 
                            yes="Green (Minimum RQI > 7)", 
                            no = ifelse(RQI.low < 7 & RQI.low > 4, yes = "yellow", no = "Red (Minimum RQI < 4)")))%>% 
  mutate(sets = factor(sets, levels = c("single", "multiple")))%>%
  dplyr::select(subject, sex, sets,benefit, csa.change, RQI.class) %>%
  spread(sets, csa.change) 



  cluster <- kmeans(temp[,c(5,6)], 2)

  temp$cluster <- cluster$cluster
    
group2 <- temp %>%
      filter(RQI.class != "Red (Minimum RQI < 4)") %>%
      ggplot(aes(single, multiple, fill = as.factor(cluster))) + 
      geom_point(shape = 21, size = 2) +
    scale_fill_manual(values = c("blue", "red"), name = "High- and low-\nrespondning cluster")
    
    

total<-mr%>%
  inner_join(rna_min)%>%
  mutate(RQI.class = ifelse(RQI.low > 7, 
                            yes="Green (Minimum RQI > 7)", 
                            no = ifelse(RQI.low < 7 & RQI.low > 4, yes = "yellow", no = "Red (Minimum RQI < 4)")))%>%
  dplyr::select(subject, sex, sets, csa.change)%>%
  spread(sets, csa.change)

incl.t <- t.test(incl$multiple, incl$single, paired=T)
total.t <- t.test(total$multiple, total$single, paired=T)

fig3+
  ylab("% Change in CSA")+xlab("Number of sets per exercise")+
  ggtitle("Figure 3. Responses to strength in relation to High- and low-integrity RNA")
```

`r subj.with.lowrqi` subjects had at least one sample with a RQI value lower than 7. These subjects do not seem to belong to either high or low responders or responders to single- or multiple-sets training (fig. 3). Excluding subjects with at least one low RQI did not change the conclusion of the study in regards of a more beneficial effect of multiple-sets vs. single-set training (mean difference between conditions `r round(incl.t$estimate, 2)` %-points 95% CI: `r round(incl.t$conf.int[1], 2)`-`r round(incl.t$conf.int[2], 2)` in dataset with lowest RQI > 7 and `r round(total.t$estimate, 2)` %-points 95% CI: `r round(total.t$conf.int[1], 2)`-`r round(total.t$conf.int[2], 2)` in the complete dataset.

### Training response clusters

Participants increasing more in response to multiple-sets compared to single-sets than the average baseline variation between legs were classified as as havining benefit of multiple-sets.

```{r,include=TRUE, warning=FALSE, echo=FALSE, message=FALSE, fig.height=4, fig.width=6}
group1 + ggtitle("Figure 4. Responders to multiple-sets") + xlab("Single-sets %-change")+
  ylab("Multiple-sets %-change")
```

Another grouping was created based on k-means clusters using multiple- and single-sets responses.

```{r,include=TRUE, warning=FALSE, echo=FALSE, message=FALSE, fig.height=4, fig.width=6}
group2 + ggtitle("Figure 5. High- vs. Low responders") + xlab("Single-sets %-change")+
  ylab("Multiple-sets %-change")
```

### RNA purity
```{r RNA purity, include=TRUE, warning=FALSE, echo=FALSE, message=FALSE, cache=TRUE}

source("./R/import_RNA_purity.R")

### Exclude subjects with at least one RNA quality below 7
rna.purity<-RNA%>%
  group_by(subject, timepoint, leg, sample.id)%>%
  summarise(RQI = mean(RQI))%>%
  inner_join(rna_purity_data)%>%
  inner_join(rna_min)%>%
  filter(RQI.low > 7)%>%
  dplyr::select(subject, timepoint, leg, sample.id, RQI, ratio.260.280, ratio.260.230)


fig4a<-rna.purity%>%
  ggplot(aes(RQI, ratio.260.280))+
  geom_point(shape=21, fill="blue", alpha=5/10, size=2)+
  ylab("Absorbance 260/280 ratio")
fig4b<-rna.purity%>%
  ggplot(aes(RQI, ratio.260.230))+
  geom_point(shape=21, fill="red", alpha=5/10, size=2)+
  ylab("Absorbance 260/230 ratio")

library(cowplot)
plot_grid(fig4a, fig4b,  labels=c("A", "B"))

purity.summary<-rna.purity%>%
  group_by()%>%
  summarise(m230 = mean(ratio.260.230),
            s230 = sd(ratio.260.230),
            m280 = mean(ratio.260.280),
            s280 = sd(ratio.260.280))



```
Average 260/280 absorbance ratio was `r round(purity.summary[1,3],2)` (SD: `r round(purity.summary[1,4],2)`) (fig 4A) and 260/230 ratio was `r round(purity.summary[1,1], 2)` (SD: `r round(purity.summary[1,2], 2)`) (fig 4B).



